%
% $Id: exception.tex,v 1.1.1.1 2004/07/12 14:07:44 cvs Exp $
%

\subsection{Exception Handling}

\subsubsection{Operation}

\subsubsection{Exception Structure Definition}

The structure definition of an exception and the ancillary structures, is as follows:

\bcode
\begin{verbatim}
 1: typedef void (*x_exception_cb)(void * arg);
 2: typedef struct x_Xcb * x_xcb;
 3:
 4: typedef struct x_Xcb {
 5:   x_xcb previous;
 6:   x_xcb next;
 7:   x_exception_cb cb;
 8:   void * arg;
 9: } x_Xcb;
10:
11: typedef struct x_Exception * x_exception;
12:
13: typedef struct x_Mutex {
14:   void * pc;
15:   void * sp;
16:   unsigned int registers[NUM_CALLEE_SAVED];
17:   x_boolean fired;
18:   x_xcb callbacks;
19:   x_Xcb Callbacks;
20: } x_Exception;
\end{verbatim}
\ecode

The relevant fields in the exception structure are the following:

\begin{itemize}
\item \txt{x\_exception$\rightarrow$pc} This field is used by the
\textsf{x\_context\_save} function to store the program counter of the
instruction following the call.
\item \txt{x\_exception$\rightarrow$sp} This field is used by the
\textsf{x\_context\_save} function to store the stack pointer at the moment
of the the call.
\item \txt{x\_exception$\rightarrow$registers} Is used to store all the
registers that are normally saved by the callee\footnote{The function that
is being called.}. This array is filled by the CPU specific
\textsf{x\_context\_save} function. Its contents are later restored when an
exception is being thrown, by the \textsf{x\_context\_restore} function. The
size of this array is CPU specific and the macro definition
\textsf{NUM\_CALLEE\_SAVED} is set in the include file of the specific CPU.
\item \txt{x\_exception$\rightarrow$fired} Is a boolean that indicates wether
an exception has been thrown or not.
\item \txt{x\_exception$\rightarrow$callbacks} Is the begin of a circular
linked list of callbacks that will be executed when an exception has been
thrown. This pointer is preset to the address of the embedded
\textsf{Callbacks} structure that is the last element of this structure.
\end{itemize}

\subsubsection{Using Native Exceptions}
